Epistemic Semantics in Guarded String Models

### 2020-09-26 tex/rdraft.tex
Edits all over in sdraft.

\newcommand{\ehc}[1]{\textit{\textcolor{red}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}
%% \newcommand{\ehc}[1]{}
%% \newcommand{\mr}[1]{}

\newcommand{\K}{\mathcal{K}}
\newcommand{\B}{\mathcal{B}}

@@

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o1  h:h \\
            event o0  t:t \\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent amy \\
            \hspace{3mm} o1  -> o1 \\
            \hspace{3mm} o0  -> o0 \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

@@

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{llll}
      \multicolumn{4}{l}{\text{events}~~e\in \mathsf{P}} \\
      p,q &\multicolumn{3}{l}{::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap} \\
      \Box_a p & \triangleq \neg\Diamond_a\neg p &
      \bullet & \triangleq \sum_{e \in \mathsf P}e \\      
      p \wedge q &  \triangleq \neg(\neg p + \neg q) &
      p \to q & \neg p + q      
      \end{array} \)
  \end{minipage}}

@@

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
\ehc{I took the `that's out of these sentences. I remember from class
  that even if they are assigned null in the PF they still exist in
  the LF. Is it okay to leave it out here?}

\subsection{Standard Epistemic Modalities}

Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$ into
Epik\footnote{The choice to refer to $\Diamond_a$ as an ``epistemic''
  primitive rather than a ``doxastic'' primitive, is that it is
  essential in constructing both the doxastic and epistemic
  operators. A ``doxastic'' moniker would suggest the
  unrepresentability of epistemic modalities}. These encodings are
shown below:

\begin{example}\label{encodings}
  \( \begin{array}[t]{ccl}
    \textsc{Belief} & \B_a~p &\triangleq \Box_a~p \\
    \textsc{Knowledge} & \K_a~p & \triangleq p \wedge \B_a~p
    \end{array}\)
\end{example}

%% As a consequence of these encodings, the standard dualities
%% hold. These are shown in \ex{modalprops}.
%% \begin{example} \label{modalprops}
%%   \(\K_a~p \to \B_a~p \qquad \B_a~p \to \neg\K_a\neg p \)
%% \end{example}

To construct different types of reasoners (e.g. accurate, inaccurate,
etc), users of Epik can construct an agents alternative relation
accordingly. Observe the discussion of modal axioms \textbf{T} and
\textbf{D} in Section~\ref{sec:axioms}.

\subsection{Guarded String Interpretation}
\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

@@

\subsection{Axiomatic Classification}
\label{sec:axioms}

To situate our logic as a modal logic, we investigate the soundness of
the standard modal axioms~\cite{????} in Epik's semantics. Some of
these standard axioms (\ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

\begin{example}\label{valid-axioms}
  \begin{enumerate}[align=left]
  \item[\textbf{N}] If $p$ is valid, then $\Box_ap$ is valid
  \item[\textbf{K}] $\Box_a(p \to q) \to \Box_a~p \to \Box_a~q$ is valid.
  \end{enumerate}
\end{example}

\begin{example}\label{conditionally-valid-axioms}
  \(\begin{array}[t]{lll}
  \textbf{T} & \Box_a p \to p & \text{if } g\mathrel{\hat{R}_a} g, \forall g \\
  \textbf{D} & \Box_a p \to \Diamond_a p & \text{if } g \in \dom(\hat{R}_a), \forall g \\
  \textbf{4} & \Box_a p \to \Box_a\Box_a p & \text{if } \hat{R}_a \text{ idempotent}
  \end{array}\)
\end{example}

The condition on \textbf{4} is essentially just a restatement of the
theorem in relational terms. The conditions on the remaining standard
modal axioms, \textbf{B} and \textbf{5}, are of a similarly trivial
flavor.

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic KAT inside the finite state
calculus. We describe our translation from Epik terms to Fst programs
here.

First, the space of worlds is a set of ordinary (as opposed to
guarded) strings.  Bit sequences (sequences of $0$'s and $1$'s) encode
atoms, as before, these alternate with event symbols to encode a
world. \ex{gs1fst} gives the encoding of worlds $w_2$ and $w_3$ from
the example. The valuation of $h$ preceeds the valuation of $t$, so
$1~0$ indicates the valuation $h\bar{t}$.

%% unnecessary detail
%% A string is a finite sequence of symbols, and {\tt 0}, {\tt 1}, {\tt
%%   u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
%% multi-character symbols that are found in implementations of the
%% finite state calculus. The multi-character symbol {\tt a0} is an
%% element of the alphabet that has no connection with the element {\tt
%%   a}.

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St | Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}
\end{tcolorbox}
\caption{\raggedright Definition in Fst of KAT product and star.
  %% Where {\tt X} and {\tt Y} are regular sets and
  %% {\tt R} and {\tt S} are regular relatiions,
  %% {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  %% {\tt X|Y} is the union of {\tt X} and {\tt Y},
  %% {\tt \~{}X} is the complement of {\tt X},
  %% {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  %% {\tt R{.}l} is the co-domain of {\tt R}, and
  %% {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

\begin{example}   \label{gs1fst}
  \renewcommand{\arraystretch}{1.2}
\begin{tabular}[t]{cl}
%% {\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
%% $w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 1 0} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.

Now, translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
ensuring to decorate the events $e$ with their compatible atoms
first. For example $a_1$ becomes $\{01a_101\}$ Fst also has built-in notions of union (\texttt{|}),
intersection (\texttt{\&}), and complement (\texttt{\~}). Translating
the the remaining operators requires more work.

Implementing fusion product using Fst's standard string concatenation
(denoted by juxatposition $\texttt{X} \texttt{Y}$) requires more
care. Naively concatenating strings with atoms (Boolean vectors) at
both ends doubles the atoms at the juncture, and does not enforce the
requisite atom equality.

To implement the fusion product, we define the binary product
operation \texttt{Cn}, which combines strings in the underlying string
algebra, remove strings with non-matching atoms, and then delete the
second of two tests to create well-formed guarded strings (see Figure
\ref{fstfigure1}). The set {\tt Wf0} describes pairs of equal atoms as
the complement of set of unequal atom pairs, implemented using Fst's
containment operator \texttt{\$}. Then. to delete the second of the
two (matching) pairs, the \texttt{Squash} operator uses Fst's rewrite
notation to delete atoms (\texttt{St}) that are preceded by another
atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used
  with relation composition and relation domain and co-domain.  They
  are restricted in that the complement and set difference for
  non-equal length relations is not defined.  The epistemic alterative
  relations that are defined in Figure -- are equal-length relations.
} The squash operator is applied via the relational composition
\texttt{.o.} and codomain \texttt{.l} operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

Kleene star is defined in a similar way (see Figure \ref{fstfigure1}),
unioning the set of atoms \texttt{St} with \texttt{Kpl} which uses
Fst's non-empty iteration operator \texttt{+} and normalizes the
result in the same way as \texttt{Cn}.

@@

\begin{hangingpar}
{\tt{define RelKst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}\\ The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}
\end{tcolorbox}
\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{fstfigure2}
\end{figure}

%% An event symbol such as $a_1$ (Amy looks at heads) is in the KAT algebra a set of
%% bare events decorated with compatible tests on each side, $\{ 10 a_1 10
%% \}$ in this case.  This is a unit set rather than a guarded string, because elements
%% of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
%% events using {\em Kst}. The operation enforces compatibility of states, so that
%% $(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
%% Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
%% KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).

In Fst, we use the Kleene closure of the concatenation product
operation on relations to lift a relation on events for an agent $a$,
$R_a$, to the corresponding relation on worlds.  First define the
concatenation product $R~S$ of two relations $R$ and $S$ to be the set
of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where $x_1 \mathrel R
y_1$, and $x_2 \mathrel S y_2$. Then $R+$ is the closure of $R$ with
respect to this operation. Figure \ref{relclosure} defines the
corresponding operation in the guarded string algebra.  The epistemic
alternative relation on worlds for an agent is then defined as the
concatenation closure of the event alternative relation for the agent.

@@

We also implement the semantics of Epik terms using lazy lists in
Haskell, rather than the direct interpretation as sets. Using lists
sidesteps checking the set invariant (elements are unique) for large
sets, such as $\bullet^*$, and laziness allows us to delay computing
these large sets until they are actually needed. To sidestep the
infiteness of models, we parameterize the interpretation function on a
positive integer $n$ and only produce guarded strings of length $n$ or
less.

%% We translate the Epik terms into a Haskell algebraic datatype that
%% represents terms with the same signature as described in Section 2. We
%% then paramaterize the interpretation function on an integer $n$ that
%% describes the maximum length of a string we will produce.

The bounded interpretation into lists of strings is very similar to
the unboundeds interpretation into sets of strings, except for the
bounds checking. The full details are shown in
Figure~\ref{fig:list-interp}. First note that when $n=0$, the
denotation is empty, denoted $[]$. Terms of the form $0$, $1$, $e$,
and $\psi$ have the same denotation as before, translated into a list
(denoted $\lfloor S\rfloor$, for a set $S$). We compute atoms using
BBDs, which concisely represent boolean functions~\cite{BBDS}.

We lift the remaining operators (except Kleene star) to their list
equivalents: union becomes list append (written $\texttt{+\!+}$);
fusion product is lifted to lists instead of sets, negation is
implemented using list difference ($\setminus$), and the modal
operator lifts the alternative relation over lists of
strings\footnote{Figure~\ref{fig:list-interp} depicts this using the
  list comprehension notation, which is analogous to set builder
  notation, except that it is written using square brackets. Element
  order is evoked by the keyword \texttt{for}, rather than using the
  unordered $\forall$.}. The only caveat to these direct
interpretations is that we restrict the operators to have size $\leq
n$, denoted as $l|_n$ for a list of guarded strings $l$.

@@

%% The lists we use here are \emph{lazy} (as opposed to \emph{strict}), which
%% broadly means that computation is delayed until the value is needed. This allows
%% us to avoid computing large, unnecessary iterations.

@@

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & ${R}_a$\\
        Bob & $e$ & ${R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
      \end{tabular}
  \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a logical language that extends the Epik term language with lambda. }
  \label{lexiconfigure}
\end{figure}

@@

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by application of semantic
translations, accompanied by beta reductions that eliminate all
lambdas in logical forms for clauses.  In consequence, the semantic
term translating a sentence is an Epik term.  Such a term designates a
set of possible words (guarded strings).  By way of example,
\exi{three}{a} is an English sentence with conjunction and several
levels of clausal embedding.  Using the grammar and parser, the
sentence is mapped to the term in \exi{three}{b}. Then \exi{three}{c}
shows a simplified logical from constructed form \exi{three}{b} using
simply logical equivalences. This term is compiled in an
implementation of the finite state calculus to a finite state machine
with six states and eight arcs, which accepts a countably infinite set
of worlds. \ehc{Is this right anymore? the formula has changed
  drastically}. In this way the methodology ``directly'' represents
the set of worlds denoted by \exi{three}{a}.

\newcommand{\HEADS}{\bullet^*h}
\newcommand{\AKWH}{\neg (\neg \HEADS \vee \Diamond_{a} \neg\HEADS) + \neg(\HEADS \wedge \Diamond_{a}\HEADS)}
\newcommand{\BKAKWH}{\neg(\neg(\AKWH) + \Diamond_{b}\neg(\AKWH))}
\newcommand{\AKBKAKWH}{\neg(\neg(\BKAKWH) + \Diamond_{a}\neg(\BKAKWH))}

\begin{examples} \label{three}
\itema It's tails and Amy knows that Bob knows that Amy knows whether it's heads.
% (_*;t) & ~ <amy> ~ ~ <bob> ~((~ (<amy> ~ (_*;h))) + ~ (<amy> (_*;h)) )
\itemb \( \displaystyle \bullet^*t \wedge \AKBKAKWH \)
\itemc \(\bullet^*t \wedge \K_{a}(\K_{b}(\K_{a} \bullet^* h + \K_{a} \neg \bullet ^* h))\)
\itemd Amy knows that it's tails.
\iteme \(
\bullet ^*t \wedge \neg \Diamond_{a} \neg \bullet^*t~~~~~~(\equiv \K_a \bullet^* t)
\)

>>>>

\newcommand{\ehc}[1]{\textit{\textcolor{green}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}

@@

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o1  h:h \\
            event o0  t:t \\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent aly \\
            \hspace{3mm} o1  -> o1 \\
            \hspace{3mm} o0  -> o0 \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

@@

\begin{figure}
  \fbox{
    \begin{minipage}{\columnwidth}
      \(\begin{array}{ll}
      \text{events} & e \in \mathsf{P} \\
      \text{states} & \sigma~\text{as in Figure 1} \\    
      p,q & ::= e\,|\,\sigma\,|\,p+q\,|\,pq\,|\,p^*\,|\,\neg p\,|\, \Diamond_ap \\
      \end{array}\)

      \vspace{1mm}
      \begin{tabular}{cl}
        $\Box_a p$ & $\triangleq \neg\Diamond_a\neg p$ \\
        \(p \wedge q\) &  \(\triangleq \neg(\neg p + \neg q)\) \\
        $\bullet$ & \( \triangleq \sum_{e \in \mathsf P}e\)
      \end{tabular}
  \end{minipage}}

@@

\begin{example}   \label{interpretation1}
\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
$\Box_a \bullet^* h$  & Amy knows that it's heads. \\
\end{tabular}
\vspace{1mm}

\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob knows that Amy knows whether \\ \hspace{3mm} it's tails.  \\
\end{tabular}
\end{example}

\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

@@

% $\tuple{K,+,\cdot,*,\bar{},0,1,{}^c,\Diamond_1,...,\Diamond_n}$

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of worlds
from the example.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

@@

\begin{hangingpar}
\begin{verbatim}
define Kst(X) St | Kpl(X);
\end{verbatim}
KAT Kleene plus and Kleene star. The Fst operation {\tt |} is union.
\end{hangingpar}
\end{tcolorbox}
\caption{\raggedright Definition in Fst of KAT product and KAT Kleene star.
  Where {\tt X} and {\tt Y} are regular sets and
  {\tt R} and {\tt S} are regular relatiions,
  {\tt X{\&}Y} is the intersection of {\tt X} and {\tt Y},
  {\tt X|Y} is the union of {\tt X} and {\tt Y},
  {\tt \~{}X} is the complement of {\tt X},
  {\tt R{.o.}S} is the composition of {\tt R} and {\tt S},
  {\tt R{.}l} is the co-domain of {\tt R}, and
  {\tt {\$}X} is the set of strings that have a substring in $X$.   
}
\label{fstfigure1}
\end{figure}

\begin{example}   \label{gs1fst}
  \renewcommand{\arraystretch}{1.2}
Worlds coded as strings
\vspace{1mm}

 \hspace{-13mm} \begin{tabular}[t]{cl}
{\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
$w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 1 0} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.  The finite state calculus has
a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has
the effect of doubling atoms at the juncture, and does not enforce
matching of atoms at the juncture.  Therefore KAT product can not be
identified with product in the finite state calculus. Instead, KAT
product and KAT Kleene star are defined operations, see Figure
\ref{fstfigure1}.  The binary product operation {\tt Cn} and the unary
Kleene star operation {\tt Kst} combine strings in the string algebra,
remove strings with non-matching atoms, and then delete the second of
two tests to create well-formed guarded strings.  Matching of atoms is
enforced with the set {\tt Wf0}, which is the set of strings that do
not contain non-matching Boolean vectors.  The containment operator
(expressed by the dollar sign) and complement (expressed by tilde) are
operators of the finite state calculus.  The set of non-matching
sequences of atoms {\tt UnequalStPair} is defined by a finite
disjunction.  Deletion is accomplished by a re-write rule in the
finite state calculus, which is a notation for defining regular
relations by contextually constrained substitutions.  In this case, is
{\tt Squash} is a regular relation that deletes an atom (a sequence of
0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state
  calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted
  in that the complement and set difference for non-equal length
  relations is not defined.  The epistemic alterative relations that
  are defined in Figure -- are equal-length relations.  }

@@

\begin{hangingpar}
{\tt{define Kst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}\\ The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}
\end{tcolorbox}
\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{fstfigure2}
\end{figure}

An event symbol such as $a_1$ (Amy looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).  In
the construction in Fst, the definition of world alternatives takes a
simple form. Where $R_a$ is the relation on decorated events for agent
$a$, the the corresponding relation on worlds in is the Kleene closure
of $R_a$.  Where $R$ and $S$ are relations, the concatenation product
of $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$,
where $\pair{x_1}{y_1}$ is in relation $R$, and $\pair{x_2}{y_2}$ is
in relation $S$.  The Kleene closure of relation $R$ is \(\cup_{n \geq
  0}R^n\), where $R^n$ is the $n$-times concatenation product of $R$
with itself (the $0$-times concatenation product is $\sem{ 1
}^{\varphi,\mathcal E}$). This is an operation in the finite
state calculus.  Figure \ref{relclosure} defines the corresponding
operation in the guarded string algebra.  The epistemic alternative
relation on worlds for an agent is then defined as the concatenation
closure of the event alternative relation for the agent.

@@

We also implement the semantics of Epik terms using lazy
lists~\cite{LazyLists} in Haskell, rather than the direct
interpretation as sets. Unfortunately, regular expressions, and hence
also Epik, can denote infinite sets of strings, for example the term
$\bullet^*$. Normally regular languages are represented using a finite
coalgebra~\cite{}. However the non-distributivity of $\diamond$
accross $;$ complicates this construction\footnote{Axiomatic and
  Coalgebraic models for Epik are open questions}. To sidestep this,
we parameterize the interpretation function on a positive integer $n$
and only produce guarded strings of length $n$ or less.

We translate the Epik terms into a Haskell algebraic datatype that
represents terms with the same signature as described in Section 2. We
then paramaterize the interpretation function on an integer $n$ that
describes the maximum length of a string we will produce.

The bounded interpretation into lists of strings is very similar to the
unbounded interpretation into sets of strings, except for the bounds
checking. The full details are shown in Figure~\ref{fig:list-interp}. First note
that when $n=0$, we simply return the empty list, denoted $[]$. Terms of the
form $0$, $1$, $e$, and $\psi$ have the same denotation as before, translated
into a list (for a set $X$, $\lfloor X \rfloor$ is a list with the same elements
as $X$ arbitrarily ordered). We compute these lists using BBDs, a standard
technique for concisely and cannoncially representing boolean functions~\cite{BBDS}.

We lift the remaining operators (except Kleene star) to their list
equivalents: union becomes list append (written $\texttt{+\!+}$);
concatenation becomes the fusion product (lifted to lists this time),
negation is implemented using list difference ($\setminus$), and the
modal operator lifts the alternative relation over lists of
strings\footnote{Figure~\ref{fig:list-interp} depicts this using the
  list comprehension notation, which is analogous to set builder
  notation, except that it is written using square brackets. Element
  order is evoked by the keyword \texttt{for}, rather than using the
  unordered $\forall$.}. The only caveat to these direct
interpretations is that we restrict the operators to have size $\leq
n$, denoted as $l|_n$ for a list of guarded strings $l$.

@@

The lists we use here are \emph{lazy} (as opposed to \emph{strict}), which
broadly means that computation is delayed until the value is needed. This allows
us to avoid computing large, unnecessary iterations.

@@

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & ${R}_a$\\
        Bob & $e$ & ${R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
      \end{tabular}
  \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a logical language that extends the Epik term language with lambda. }
  \label{lexiconfigure}
\end{figure}

@@

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by application of semantic
translations, accompanied by beta reductions that eliminate all
lambdas in logical forms for clauses.  In consequence, the semantic
term translating a sentence is a term of the logical language.  Such a
term designates a set of possible words (guarded strings) in the
possible worlds model determined by an Epik specification.  By way of
example, \exi{three}{a} is an English sentence with 
conjunction and several levels of clausal embedding.  Using the
grammar and parser, the sentence is mapped to the term in
\exi{three}{b}. The negations are propositional. The double
negation, while logically redundant, comes syntactically from the grammar. This term
is compiled in an implementation of the finite state calculus to a
finite state machine with six states and eight arcs, which accepts a
countably infinite set of worlds.  In this way the methodology
``directly'' represents the set of worlds denoted by \exi{three}{a}.

\begin{examples} \label{three}
\itema Its tails and Amy knows that Bob knows that Amy knows whether its heads.
% (_*;t) & ~ <amy> ~ ~ <bob> ~((~ (<amy> ~ (_*;h))) + ~ (<amy> (_*;h)) )
\itemb \( \displaystyle
\bullet^*t \wedge \neg \Diamond_{\text{Amy}} \neg \neg \Diamond_{\text{Bob}} \neg(
\neg \Diamond_{\text{Amy}} \neg \bullet^*h 
+
\neg \Diamond_{\text{Amy}} \bullet^*h 
)
\)
\end{examples}
Sentence \exi{entail}{a} is assigned the logical form \exi{entail}{b} by the grammar.
Logical relations between propositions are checked in the finite state calculus
by checking set-theoretic relations between sets of worlds.
In this case, the propositions $p_{11}$ (from \ex{three}) and
$p_{12}$ (from \ex{entail}) have an empty
symmetric difference 
$(p_{11} - p_{12}) + (p_{12} - p_{11})$, and so sentences
\exi{three}{a} and \exi{entail}{b} are equivalent, in the sense that they
denote the same set of worlds.  This relation is assessed computationally
in the possible worlds model.
